### 2022-06-12 paper.tex
a few changes to the benchmark section

\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\urlstyle{same} % no tt font for URLs
\usepackage{natbib}
\bibliographystyle{genome_research}
%\setcitestyle{aysep={}} 
\usepackage[dvipsnames]{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!50!black}
}
\renewcommand{\dbltopfraction}{0.7}
\renewcommand{\textfraction}{0.2}

@@

@@

CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to spontaneous deamination), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription. The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citet{cedar2022role}. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citet{argelaguet2019gastru}.

@@

The standard approach to analyse scBS data is based on methodology developed for the analysis of single-cell RNA-Seq data. Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.

@@

\begin{figure*}
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the scbs package}
    }
    \label{workflow}
\end{figure*}

@@

In brief, we store methylation of each chromosome in a binary matrix where each column represents a cell and each row represents a base pair. Since the vast majority of values in this matrix are missing due to the sparsity of scBS data, we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format. This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.

@@

After quality control, the user has access to the previously described genome-wide VMR detection approach via \texttt{scbs scan}. This produces a BED-file that lists the genome coordinates of VMRs. To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED-file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells. The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells. We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells. The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

\section{Benchmarks}
To assess performance of our newly devised methods, we obtained two previously published scBS data sets and re-analysed them with \texttt{scbs}.

The first data set is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}. In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells. They calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development. The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes. We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.

@@

To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}).
Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we obtained a second data set from \citet{luo2017single}. This data set consists of over three thousand single-cell CpG methylomes from murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.

To assess performance of each combination of methods, we used a nearest neighbor count score that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space (see Methods).
The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).
Most of this gain in performance can be attributed to the use of VMRs over genomic tiles. The use of mean shrunken residuals over methylation fractions provides an additional small improvement.
Of note, our results show that even few VMRs, e.g. the 1000 VMRs that have read coverage in the highest number of cells, already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
% why not just plot PC1 vs. PC2 here?

Since some scBS protocols are based on the use of 96-well-plates, and since scBS is costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
In this smaller data set, the performance gains were even more pronounced, indicating that our methods help extract information even from challenging data sets (Figure \ref{figure:score}D).

\begin{figure*}
    \begin{center}
    A\hspace{.35\textwidth}B\\
    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
    C\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
    D\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs using 1000 high-coverage intervals of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs using 1000 intervals of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right). Data from \citet{luo2017single}.}
    \label{figure:score}
    \end{center}
\end{figure*}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\urlstyle{same} % no tt font for URLs
\usepackage{natbib}
\bibliographystyle{genome_research}
%\setcitestyle{aysep={}} 
\usepackage[dvipsnames]{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!50!black}
}
\renewcommand{\dbltopfraction}{0.9}
\renewcommand{\textfraction}{0.15}

@@

\newcommand{\todo}[1]{[\textcolor{orange}{#1}]}

@@

CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to spontaneous deamination), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription. The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citep{cedar2022role}. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citep{argelaguet2019gastru}.

@@

The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data. Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.

@@

\begin{figure*}
    \begin{center}
        \includegraphics[width=.85\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the scbs package}
    }
    \label{workflow}
\end{figure*}

@@

In brief, we store methylation of each chromosome in a matrix where each column represents a cell and each row represents a base pair. Methylated and unmethylated sites are coded as $+1$ and $-1$, respectivelty. Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format. This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.

@@

After quality control, the user has access to the genome-wide VMR detection approach described above via \texttt{scbs scan}. This produces a BED file that lists the genome coordinates of VMRs. To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells. The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells. We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells. The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

\section{Benchmarks}
To assess performance of our methods, we obtained two previously published scBS-Seq data sets and re-analysed them with \texttt{scbs}.

\subsection{Correlation with gene expression}

The first data set is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}. In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells. In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development. The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes. We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.

@@

To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.

\subsection{Identification of cellular sub-types}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used the data set from \citet{luo2017single}. This data set consists of over three thousand \todo{Give exact number.} single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. \todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kep only the top 1000 regions.

To assess performance of each combination of methods, we used a nearest neighbor count score that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space (see Methods).
The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).
Most of this gain in performance can be attributed to the use of VMRs over genomic tiles. The use of mean shrunken residuals over methylation fractions provides an additional small improvement.

Since some scBS protocols are based on the use of 96-well-plates, and since scBS is costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{figure:score}B).

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).

\begin{figure*}
    \begin{center}
    A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
    \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
    C\hspace*{.7\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
    D\hspace*{.7\textwidth}~\\
    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right). Data from \citet{luo2017single}.}
    \label{figure:score}
    \end{center}
\end{figure*}
